---
title: Introduction to dataframes in R
author: Mitul Patel
date: '2020-08-14'
slug: introduction-to-dataframes-in-r.en-us
categories:
  - R basics
tags:
  - dataframe
  - R
keywords:
  - tech
---

In this second blog post in the series of data science in R, I will introduce the basics of data frames. The credit for all the material in this and the future posts goes to [DataCamp](https://datacamp.com/).
<!--more-->

In this post I will introduce data frames in R. The credit for all the material in this and the future posts goes to [DataCamp](https://datacamp.com/).

# Dataframes

Data frames are bascially created from vectors. It consists of rows (i.e observations) and columns (i.e variables or features). Data frames  holds a different type of data in each column. We can create a dataframe in R using `data.frame()`. Let's create 3 variables and use them to create a data frame and assign it to a variable `cash` as follows

```{r}
company <- c("A", "A", "A", "B", "B", "B", "B")
cash_flow <- c(1000, 4000, 550, 1500, 1100, 750, 6000)
year <- c(1, 3, 4, 1, 2, 4, 5)
cash <- data.frame(
    company = company,
    cash_flow = cash_flow, 
    year = year)
cash
```

We can look at the first few rows and/or tails and/or structure of the data frame using `head()` and/or `tail()` and/or `str()` respectively. 
```{r}
head(cash, n=2)
tail(cash, n=2)
str(cash)
```
`str` is useful if you want to check the data types of each columns in your data frame.

Just like with vectors and matrices, we can extract or access certain columns or rows of the data frame using `[row index, col index/col name]`, where you just pass in row index and column index or column name.  Let's look at how we do this with the `cash` data frame:

```{r, collapse=TRUE}
#thrid row and 2nd col
cash[3,2]  

#fifth row and cash_flow col - notice we used column name here 
cash[5,"cash_flow"] 
```

To extract a specific column, there is shortcut that you can use, `$`
```{r}
# extract "year" column using "$"
cash$year

cash$cash_flow *2 # manipulate data
cash
```

We can also subset a data frame using `subset()` by passing the first argument as data and the second argument as a filter. In the examples below, we subset the data frame to only `cash` corresponding to company `A`and rows that have cash flows due in 1 year:
```{r}
subset(cash, company == "A") # cash flow for only company A
subset(cash, year == 1) # row having cash flow due in a year
```

We can add a new column using `data_frame$new_column`. Let's add `quarter_cash` column to our data frame by transforming existing columns
```{r}
# Quarter cash flow scenario
cash$quarter_cash <- cash$cash_flow * .25

# Double year scenario
cash$double_year <- cash$year * 2
cash

cash$company <- NULL # delete column
cash
```

There is alot that exist about data frames which I haven't covered in this post, as the aim of these posts to keep it short. Hopefully this will atleast get you started with data frames in R. 

