---
title: "Working with categorical data in R using Tidyverse"
author: "Mitul Patel"
date: "2020-07-21"
slug: categorical data
tags: ["R", "categorical", "tidyverse/tidyr", "stringr", "dplyr", "stringr"]
categories: ["Categorical data"]
output:
  blogdown::html_page:
    highlight: tango
---


<style>
body {text-align: justify};
</style>

The libraries we will be using in this blog are 
```{r, message=FALSE}
library(data.table)
library(dplyr)
library(tidyverse)
```

## Into to factors variables

Here are the two sample datasets we will be using, one of the dataset is the kaggle survery data
```{r}
flying_etiquette <- fread("~/Desktop/R_tutorials/data/flying-etiquette.csv") %>% 
    as_tibble()
str(flying_etiquette)

kaggle_sur <- fread("~/Desktop/R_tutorials/data/kaggle_sur.csv") %>%
    as_tibble()
str(kaggle_sur)
```

To check whether a particular variable is a factor or not, we simply pass the data and the variable to `is.factor()` and we get a boolean as an output

```{r}
is.factor(kaggle_sur$CurrentJobTitleSelect)
```
We get `FALSE` which means we will need to convert this to a factor. 

You are probabaly aware of `mutate()` and `summarise()` in **dply**, here we look at slight modified version of these `mutate_if()`, `summarise_if`, `mutate_all()` and `summarise_all`. `mutate_if()` and `summarise_if` apply their second argument, a function, to all columns where the first argument is true, and `mutate_all()` and `summarise_all()` take one argument, a function, and apply it to all columns. Let's convert all the columns that are characters (note first argument has to be true) to factors

```{r}
responses_as_factors <- kaggle_sur %>%
    mutate_if(is.character, as.factor)
str(responses_as_factors)
```

Now we can find the number of levels in each column using
```{r, collapse=TRUE}
number_of_levels <- responses_as_factors %>%
    summarise_all(nlevels) %>%
    gather(variable, num_levels) # just to format the data from wide to long
number_of_levels
```

we can also look at 3 rows with the highest number of levels
```{r}
number_of_levels %>% top_n(3, num_levels)
```

If we want to look at the number of level of a specific variable we can do
```{r}

number_of_levels %>%
    filter(variable == "CurrentJobTitleSelect") %>%
    pull(num_levels) # extracts the value of that column/variable
```
or we can even see what levels there are in this column
```{r}
responses_as_factors %>%
    pull(CurrentJobTitleSelect) %>%
    levels()
```

and we can plot these levels as a bar plot since the column is categorical 
```{r}
ggplot(kaggle_sur, aes(CurrentJobTitleSelect)) +
  geom_bar() +
  coord_flip() +
  xlab("Current Job Title")
```
note, how the plot is unordered and sometimes hard to compare each of the levels. We can make the bar plot ordered using `fct_infreq()` from the **forcats** package. Just an fyi, anything that starts with **fct** comes from the **forcats** package

```{r}
ggplot(kaggle_sur, aes(fct_rev(fct_infreq(CurrentJobTitleSelect)))) +
  geom_bar() +
  coord_flip() +
  xlab("Current Job Title")
```
also here I've used `fct_rev()` to reverse the order of the bar plot, so from top to bottom it goes __highest__ count to __lowest__.

We can look at the mean age of the people with each of these job title. Here we make use of `fct_reorder()` to order one varaiable based on another
```{r, warning=FALSE, message=FALSE}
kaggle_sur %>% 
  filter(!is.na(Age) & !is.na(CurrentJobTitleSelect)) %>% # don't include NA's 
  group_by(CurrentJobTitleSelect) %>% 
  summarise(mean_age = mean(Age)) %>% 
  mutate(CurrentJobTitleSelect =
           fct_reorder(CurrentJobTitleSelect,mean_age)) %>% # reorder job title based on mean age column
  ggplot(aes(x = CurrentJobTitleSelect, y = mean_age)) + 
  geom_point() +
  coord_flip()
  
```

## Manipulating factor variables
```{r}
levels(responses_as_factors$WorkInternalVsExternalTools)
mc_responses_reordered <- responses_as_factors %>%
    mutate(WorkInternalVsExternalTools =
             fct_relevel(WorkInternalVsExternalTools,
                         "Entirely internal", "More internal
                         than external", "Approximately half
                         internal and half external", "More
                         external than internal", "Entirely
                         external", "Do not know"))
ggplot(mc_responses_reordered, aes(WorkInternalVsExternalTools)) + 
    geom_bar() + 
    coord_flip()

levels(responses_as_factors$FormalEducation)
responses_as_factors %>%
    # Move "I did not complete any formal education past high school" and "Some college/university study without earning a bachelor's degree" to the front
    mutate(
    FormalEducation = fct_relevel(FormalEducation, 
                                  "I did not complete any formal education past high school", 
                                  "Some college/university study without earning a bachelor's degree")) %>%
    # Move "I prefer not to answer" to be the last level.
    mutate(
      FormalEducation = fct_relevel(FormalEducation, 
                                    "I prefer not to answer" , 
                                    after = Inf)) %>%
    # Move "Doctoral degree" to be after the 5th level
    mutate(
      FormalEducation = fct_relevel(FormalEducation, 
                                    "Doctoral degree", 
                                    after = 5)) %>%
    # Examine the new level order
    pull(FormalEducation) %>%
    levels()

ggplot(responses_as_factors, aes(FormalEducation)) + 
    geom_bar()

responses_as_factors %>%
    # rename the appropriate levels to "High school" and "Some college"
    mutate(FormalEducation = fct_recode(FormalEducation,
    "High school" = "I did not complete any formal education past high school", 
    "Some college" = "Some college/university study without earning a bachelor's degree",
    "No answer" = "I prefer not to answer")) %>%
    # make a bar plot of FormalEducation
    ggplot(aes(x = FormalEducation)) + 
    geom_bar()

responses_as_factors %>%
    # Create new variable, grouped_titles, by collapsing levels in CurrentJobTitleSelect
    mutate(grouped_titles = fct_collapse(CurrentJobTitleSelect, 
        "Computer Scientist" = c("Programmer", "Software Developer/Software Engineer"), 
        "Researcher" = "Scientist/Researcher", 
        "Data Analyst/Scientist/Engineer" = c("DBA/Database Engineer", "Data Scientist", 
                                          "Business Analyst", "Data Analyst", 
                                          "Data Miner", "Predictive Modeler"))) %>%
    # Keep all the new titles and turn every other title into "Other"
    mutate(grouped_titles = fct_other(grouped_titles, 
                                      keep = c("Computer Scientist", 
                                               "Researcher", 
                                               "Data Analyst/Scientist/Engineer"))) %>% 
    # Get a count of the grouped titles
    count(grouped_titles)

responses_as_factors %>%
  # remove NAs of MLMethodNextYearSelect
  filter(!is.na(MLMethodNextYearSelect)) %>%
  # create ml_method, which lumps all those with less than 5% of people into "Other"
  mutate(ml_method = fct_lump(MLMethodNextYearSelect, prop = .05)) %>%
  # count the frequency of your new variable, sorted in descending order
  count(ml_method, sort = TRUE)

responses_as_factors %>%
  # remove NAs 
  filter(!is.na(MLMethodNextYearSelect)) %>%
  # create ml_method, retaining the 5 most common methods and renaming others "other method" 
  mutate(ml_method = fct_lump(MLMethodNextYearSelect, n = 5, other_level = "other method")) %>%
  # count the frequency of your new variable, sorted in descending order
  count(ml_method, sort = TRUE)
```
recode - to change bunch of values in a col to something else. Remember that when recoding numeric variables, you need to put the old value in backticks. .default option see Working with Data in the Tidyverse n parse_number.

## Creating factors variable
```{r}
learning_platform_usefulness <- responses_as_factors %>%
  # select columns with LearningPlatformUsefulness in title
  select(contains("LearningPlatformUsefulness")) %>%
  # change data from wide to long
  gather(learning_platform, usefulness) %>%
  # remove rows where usefulness is NA
  filter(!is.na(usefulness)) %>%
  # remove "LearningPlatformUsefulness" from each string in  c 
  mutate(learning_platform = str_remove(learning_platform,"LearningPlatformUsefulness"))

perc_useful_platform <- learning_platform_usefulness %>% 
  count(learning_platform, usefulness) %>%
  add_count(learning_platform, wt = n, name = "total") %>% 
  mutate(perc = n / total)
ggplot(perc_useful_platform, aes(x = usefulness, y = perc, group = learning_platform)) + 
  geom_line() + 
  facet_wrap(~ learning_platform)
```

```{r}
usefulness_by_platform <- learning_platform_usefulness %>%
    # If usefulness is "Not Useful", make 0, else 1 
    mutate(usefulness = if_else(usefulness == "Not Useful", 0, 1)) %>%
    # Group by learning platform 
    group_by(learning_platform) %>%
    # Summarize the mean usefulness for each platform
    summarize(avg_usefulness = mean(usefulness))

ggplot(usefulness_by_platform, aes(x = learning_platform, y= avg_usefulness)) + 
  geom_point() +
  theme(axis.text.x =element_text(angle = 90, hjust = 1)) +
  labs(x= "Learning Platform", y= "Percent finding at least somewhat useful") + 
  scale_y_continuous(labels = scales::percent)

usefulness_by_platform %>%
	# reorder learning_platform by avg_usefulness
	mutate(learning_platform = fct_reorder(learning_platform, avg_usefulness)) %>%
	# reverse the order of learning_platform
	mutate(learning_platform = fct_rev(learning_platform)) %>%
	ggplot(aes(x = learning_platform, y = avg_usefulness)) + 
	geom_point() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    labs(x = "Learning Platform", y = "Percent finding at least somewhat useful") + 
    scale_y_continuous(labels = scales::percent)
  
```

```{r}
responses_as_factors %>%
    # Filter for rows where Age is between 10 and 90
    filter(between(Age, 10, 90)) %>%
    # Create the generation variable based on age
    mutate(generation = case_when(
    between(Age, 10, 22) ~ "Gen Z", 
    between(Age, 23, 37) ~ "Gen Y", 
    between(Age, 38, 52) ~ "Gen X", 
    between(Age, 53, 71) ~ "Baby Boomer", 
    between(Age, 72, 90) ~ "Silent"
    )) %>%
    # Get a count of how many answers in each generation
    count(generation)
```
```{r}
multiple_choice_responses <-  responses_as_factors %>% 
  mutate(JobSatisfaction = recode(JobSatisfaction, "I prefer not to share" = 'NA')) %>% 
  separate(JobSatisfaction, into = ("JobSatisfaction"), sep = "-") 
multiple_choice_responses$JobSatisfaction = as.numeric(multiple_choice_responses$JobSatisfaction)
multiple_choice_responses %>%
    # Filter out people who selected Data Scientist as their Job Title
    filter(CurrentJobTitleSelect != "Data Scientist") %>%
    # Create a new variable, job_identity
    mutate(job_identity = case_when(
        CurrentJobTitleSelect == "Data Analyst" & DataScienceIdentitySelect == "Yes" ~ "DS analysts", 
        CurrentJobTitleSelect == "Data Analyst" & DataScienceIdentitySelect %in% c("No", "Sort of (Explain more)") ~ "NDS analyst", 
        CurrentJobTitleSelect != "Data Analyst" & DataScienceIdentitySelect == "Yes" ~ "DS non-analysts", 
        TRUE ~ "NDS non analysts")) %>%
    # Get the average job satisfaction by job_identity, removing NAs
    group_by(job_identity) %>%
    summarize(avg_js = mean(JobSatisfaction, na.rm=T))
```

# Case study
```{r}
gathered_data <- flying_etiquette %>%
    mutate_if(is.character, as.factor) %>%
    filter(`How often do you travel by plane?` != "Never") %>%
    # Select columns containing "rude"
    select(contains('rude')) %>%
    # Change format from wide to long
    gather(response_var, value)
```

Expression | Does this
-----------|--------------
.	       |  matches any character
*	       |  zero or more times
           |

```{r}
gathered_data %>%
    # Remove everything before and including "rude to " (with that space at the end!)
    mutate(response_var = str_remove(response_var, ".*rude to ")) %>%
    # Remove "on a plane"
    mutate(response_var = str_remove(response_var, "on a plane"))
```


```{r}
rude_behaviors <- gathered_data %>%
    mutate(response_var = str_replace(response_var, '.*rude to ', '')) %>%
    mutate(response_var = str_replace(response_var, 'on a plane', '')) %>%
	# Remove rows that are NA in the value column
	filter(!is.na(value)) %>%
    mutate(rude = if_else(value %in% c("No, not rude at all", "No, not at all rude"), 0, 1)) %>%
    # Group by response_var
    group_by(response_var) %>%
    # Create perc_rude, the percent considering each behavior rude
    summarise(perc_rude = mean(rude))

rude_behaviors <- rude_behaviors %>% 
    mutate(response_var = fct_recode(response_var,
                                     "close to family?" = "ask someone to switch seats with you in order to be closer to family?",
                                     "close to friends?" = "ask someone to switch seats with you in order to be closer to friends?",
                                     "talk to stranger?"="say more than a few words tothe stranger sitting next to you ?",
                                     "wake passenger for bathroom?"="wake a passenger up if you are trying to go to the bathroom?",
                                     "wake passenger for walk"="wake a passenger up if you are trying to walk around?"))

rude_behaviors
```

```{r}
initial_plot <- rude_behaviors %>%
	# reorder response_var by perc_rude
	mutate(response_var = fct_reorder(response_var, perc_rude)) %>%
	# make a bar plot of perc_rude by response_var
	ggplot(aes(x = response_var, y = perc_rude)) + 
    geom_col()


# View your plot
initial_plot

titled_plot <- initial_plot + 
    # Add the title, subtitle, and caption
    labs(title = "Hell Is Other People In A Pressurized Metal Tube",
         subtitle = "% of 874 air-passenger respondents",# who said action is very or somewhat rude",
         caption = "Source: SurveyMonkey Audience", 
         # Remove the x- and y-axis labels
         x = "", 
         y = "") 

titled_plot

flipped_plot <- titled_plot + 
    # Flip the axes
    coord_flip() + 
    # Remove the x-axis ticks and labels
    theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank())
flipped_plot
```

```{r}
library(scales)
flipped_plot + 
    # Apply percent() to perc_rude to label above the bar with the perc value
    geom_text(aes(label = percent(perc_rude), 
                  y = perc_rude + .045), 
              position = position_dodge(0.9),
              vjust = 0.8)

```

