---
title: "Introduction to R basics"
author: "Mitul Patel"
date: "2020-07-21"
slug: R basics
tags: ["R", "vectors", "matrices", "dataframe", "lists", "factors"]
categories: ["R basics"]
output:
  blogdown::html_page:
    highlight: tango
---

<style>
body {text-align: justify};
</style>

In this post I will introduce the basics of R such as vectors and matrices, dataframes, lists, factors and some financial concepts. This is a first in a series of future posts to data science in financial industry using R. The credit for all the material in this and the future posts goes to [DataCamp](https://datacamp.com/).
<!--more-->

In this post I will introduce the basics of R such as vectors and matrices, dataframes, lists, factors and some financial concepts. This is a first in a series of future posts to data science in financial industry using R. The credit for all the material in this and the future posts goes to [DataCamp](https://datacamp.com/).

We can select and subset anything using [] where the first argument is always row and second is always column i.e [row, column]. 

# Vectors and matrices
creating a vector
```{r}
# numeric type
ibm_stock <- c(159.82, 160.02, 159.84)

# character type
finance <- c("stocks", "bonds",  "investments")

# logical 
logic <- c(TRUE, FALSE, TRUE)
```
Note: a vector can only be composed of one data type, otherwise the lower ranking type will be coerced into the higher ranking type. The hierarchy for coercion is:
    logical < integer < numeric < character

```{r}
a <- c(1L , "I am a character")
typeof(a)

b <- c(TRUE, "Hello")
typeof(b)

c <- c(FALSE, 2)
typeof(c)
```
In finance, the weighted average is an important calculation. It allows you to calculate your portfolio return over a time period. This is how we can use vectors to compute weighted average

Lets say you have 40% of your cash in Apple stock, and 60% of your cash in IBM stock. If, in January, Apple earned 5% and IBM earned 7%, then your total portfolio return is.... 
   return_apple * weighted_avg_apple + return_IBM + weighted_avg_IBM
```{r}  
returns <-  c(5, 7)
weights <- c(0.4, 0.6)
names(returns) <-  c("apple", "ibm")
names(weights) <-  c("apple", "ibm")
returns
weights

portfolio_return <- sum(returns * weights)
portfolio_return
```

accessing vectors
```{r}
return <- c(5, 2, 3, 7, 8, 3, 5, 9, 1, 4, 6, 3)
return[2] #2nd element
return[3:6] #3rd to 6th element
return[-1] # omit first element

# if we have col names 
names(return) <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
return[c("Feb", "Mar")]
```
creating matrices - we can create matrices from vectors

```{r}
v <- c(1, 2, 3, 4)
matrix(v, nrow = 2, ncol = 2) # notice the matrix is filled by col by default

matrix(v, nrow = 2, ncol = 2, byrow = T) # we can specify byrow = TRUE if we want matrix to be filled by row
```
We can combine multiple vectors to create matrix as follows
```{r}
apple <- c(109.49, 109.90, 109.11, 109.95, 111.03)
ibm <- c(159.82, 160.02, 159.84, 160.35, 164.79)
microsoft  <-  c(59.20, 59.25, 60.22, 59.95, 61.37)
cbind(apple, ibm) # column bind
rbind(apple, ibm) # row bind
```
If we want to know a relationship between say two stocks. We can either plot a scatter plot or we can compute correlation matrix. Correlation is a measure of association between two things, here, stock prices, and is represented by a number from -1 to 1. Few points to note:

  * 1 represents perfect positive correlation, 
  * -1 represents perfect negative correlation
  * 0 correlation means that the stocks move independently of each other. Correlation is a common metric in finance, and it is useful to know how to calculate it in R.
```{r}
cor(apple, ibm) # value close to 1 means that stocks of apple and ibm move closely together 
stocks <- cbind(apple, ibm, microsoft)
cor(stocks) #  returned a matrix that shows all of the possible pairwise correlations
```

Matrix subsetting
```{r}
stocks[2, ] # select 2nd row and all columns
stocks[, 2] # select all rows and 2nd column
stocks[c(1,2), ] # select 1st, 2nd row and all columns
stocks[, c(1,2)] # select all rows and 1st, 2nd column
stocks[c(1,2), c(1,2)] # select 1st, 2nd rows and columns
```

# Dataframes
creating dataframe
```{r}
company <- c("A", "A", "A", "B", "B", "B", "B")
cash_flow <- c(1000, 4000, 550, 1500, 1100, 750, 6000)
year <- c(1, 3, 4, 1, 2, 4, 5)
cash <- data.frame(
    company = company,
    cash_flow = cash_flow, 
    year = year)
cash
```
access and subset dataframe
```{r, collapse=TRUE}
cash[3,2] #thrid row and 2nd col 
cash[5,"year"] #fifth row and year col

# can access cols like this as well
cash$year

cash$cash_flow *2 # manipulate data
cash$company <- NULL # delete col
cash

cash <- data.frame(
    company = company,
    cash_flow = cash_flow, 
    year = year)
subset(cash, company == "B") # subsetting dataframe
subset(cash, year == 1)
```
adding new cols
```{r}
# Quarter cash flow scenario
cash$quarter_cash <- cash$cash_flow * .25

# Double year scenario
cash$double_year <- cash$year * 2
cash
```

present_value <- cash_flow * (1 + interest / 100) ^ -year
```{r}
# Present value of $4000, in 3 years, at 5%
present_value_4k <- 4000 * (1.05) ^ -3

# Present value of all cash flows
cash$present_value <- cash$cash_flow * (1.05) ^ -cash$year
cash

# Total present value of cash
total_pv <- sum(cash$present_value)

# Company B information
cash_B <- subset(cash, company == "B")

# Total present value of cash_B
total_pv_B <- sum(cash_B$present_value)
total_pv_B
```


# Factors
creating factors
```{r, collapse=TRUE}
credit_rating <- c("BB", "AAA", "AA", "CCC", "AA", "AAA", "B", "BB")
credit_factor <- factor(credit_rating)

# factor levels
levels(credit_factor)
levels(credit_factor) <- c("2A", "3A", "1B", "2B", "3C")
credit_factor
```

 bucket numerical variable into factors
```{r}
AAA_rank <- as.numeric(c("31", "48", "100", "53", "85", "73", "62", "74", "42", "38", "97", "61", "48", "86", "44", "9", "43", "18", "62", "38", "23", "37", "54", "80", "78", "93", "47", "100", "22", "22", "18", "26", "81", "17", "98", "4", "83", "5", "6", "52", "29", "44", "50", "2", "25", "19", "15", "42", "30", "27"))

AAA_factor <- cut(x = AAA_rank, breaks = c(0, 25, 50, 75, 100))
levels(AAA_factor) <- c("low", "medium", "high", "very_high")
plot(AAA_factor)
```

ordering factors
```{r, collapse=TRUE}
unique(credit_rating)
credit_factor_ordered <- factor(credit_rating, ordered = TRUE, levels = c('AAA','AA','BB','B','CCC'))
plot(credit_factor_ordered)
```

subsetting a factor
```{r, collapse=TRUE}
# Remove the A bonds at positions 3 and 7. Don't drop the A level.
keep_level <- credit_factor[-c(3,7)] # look at levels

# Plot keep_level
plot(keep_level)

# Remove the A bonds at positions 3 and 7. Drop the A level.
drop_level <- credit_factor[-c(3,7), drop = TRUE]

# Plot drop_level
plot(drop_level)
```
stringsAsFactors - when creating data frames, R is to convert all characters into factors. This causes problems at times, so we can do the following in order to avoid the problem. 
```{r}
credit_rating <- c("AAA", "A", "BB")
bond_owners <- c("Dan", "Tom", "Joe")
bonds <- data.frame(credit_rating, bond_owners, stringsAsFactors = FALSE)

# Then convert the relevant col to factor
bonds$credit_factor <- factor(bonds$credit_rating, ordered = TRUE, levels = c("AAA","A","BB"))
```


# Lists
creating lists
```{r, collapse=TRUE}

name <- "Apple and IBM"
apple <- c(109.49, 109.90, 109.11, 109.95, 111.03)
ibm <- c(159.82, 160.02, 159.84, 160.35, 164.79)
cor_matrix <- cor(cbind(apple, ibm))
portfolio <- list(name, apple, ibm, cor_matrix)
portfolio
```


named lists
```{r}
names(portfolio) <- c("portfolio_name", "apple", "ibm", "correlation")
``` 

access lists
```{r, collapse=TRUE}
portfolio[c(2,3)]
portfolio$correlation
```  

adding to lists
```{r}
portfolio$weight <- c(apple = 0.20, ibm = 0.80) # weights to portfolio
portfolio
```

removing from lists
```{r}
portfolio$ibm <- NULL
portfolio
```

split into two data frames
```{r, collapse=TRUE}
cash <- data.frame(
    company = rep(c("A","B"),each = 2),
    cash_flow = c(1000, 4000, 200, 500),
    year = c(1,3, 1,2)
)

grouping <- cash$company
split_cash <- split(cash, grouping)

original_cash <- unsplit(split_cash, grouping) # back to original
original_cash
```

Split-Apply-Combine
```{r, collapse=TRUE}
split_cash$B$cash_flow
split_cash$B$cash_flow <- 0
cash_no_A <- unsplit(split_cash, grouping)
cash_no_A
```

Attributes
```{r, collapse=TRUE}
my_matrix <- matrix(c(1,2,3,4,5,6), nrow = 2, ncol = 3)
rownames(my_matrix) <- c("Row1", "Row2")
colnames(my_matrix) <- c("Col1", "Col2", "Col3")

my_factor <- factor(c("A", "A", "B"), ordered = T, levels = c("A", "B"))

attributes(my_matrix)
attr(my_matrix, which = "dim")
attributes(my_factor)
```

