---
title: "Introduction to vectors and matrices in R"
author: "Mitul Patel"
date: "2020-07-21"
slug: R basics
tags: ["R", "vectors", "matrices", "dataframe", "lists", "factors"]
categories: ["R basics"]
output:
  blogdown::html_page:
    highlight: tango
---

<style>
body {text-align: justify};
</style>

In this series of blog posts to data science in R, I will introduce the basics of R such as vectors and matrices. The credit for all the material in this and the future posts goes to [DataCamp](https://datacamp.com/).
<!--more-->

In this post I will introduce the basics of R such as vectors and matrices. The credit for all the material in this and the future posts goes to [DataCamp](https://datacamp.com/).

A basic concept in all programming languages (as well as Math) is a **Variable**. A variable is simply a name given to something that stores a value or something to which we assign a value. For instance we can assign `4` to a variable `a` i.e `a = 4`. In R, we can do this using the following command 
```{r, collapse=TRUE}
a <- 4
a

b <- 'hello'
b

c <- TRUE
c
```
We can then perform various mathematical operations with the variable, which I won't go into details here.  

Before moving on to vectors, we need to know about data types in R. 

# Data types 

R works with numerous data types. The following are some of the most basic types you will across in R (as well as in other programming languages): 

  * **numeric**: are decimal numbers like 4.5. A special type
  of numeric is an integer, which is a numeric without a decimal
  piece. Integers must be specified like 4L.
  * **Logicals**: are the boolean values TRUE and FALSE. Capital
  letters are important here; true and false are not valid.
  * **Characters**: are text values like "hello world".
  creating a vector.

You can check the __data type__ of a variable using `class()` function:

```{r, collapse=TRUE}
numeric_type <- 20
character_type <- "hello"
logical_type <- FALSE

# Check which type these variables have:
class(numeric_type)
class(character_type)
class(logical_type)
```

Now that we know how to create variable and check its data type. Let's talk about vectors. 

# Vectors 

A vector is simply a one dimensional array i.e an object to store data. We can create a vector in R using a __combine function__ `c()` and assign it to a variable. 

```{r, collapse=TRUE}
# numeric 
ibm_stock <- c(159.82, 160.02, 159.84)
ibm_stock

# character
finance <- c("stocks", "bonds",  "investments")
finance

# boolean
logic <- c(TRUE, FALSE, TRUE)
logic

# we can also check the class of each of these vectors 
class(ibm_stock)
class(finance)
class(logic)
```
Note: a vector can only be composed of one data type, otherwise the lower ranking type will be coerced into the higher ranking type. The hierarchy for coercion is: 

> logical < integer < numeric < character 

```{r, collapse=TRUE}
a <- c(1L , "I am a character")
typeof(a)

b <- c(TRUE, "Hello")
typeof(b)

c <- c(FALSE, 2)
typeof(c)
```

We can access elements (also known as vector subsetting) from vectors using `[]`. In R, the indexing starts at 1 rather than 0 such as in Python. Let's see some examples of these,
```{r, collapse=TRUE}
a <- c(5, 2, 3, 7, 8, 3, 5, 9, 1, 4, 6, 3)
a[2] #2nd element 
a[3:6] #3rd to 6th element
a[-1] # omit first element
```


# Matrices

Matrices are similar to vectors, except they are in 2 dimensions. We can create matrices from vectors using `matrix()`and by specifying the number of rows and columns. Let's create a 2 by 2 matrix,

```{r, collapse=TRUE}
v <- c(1, 2, 3, 4)
v
matrix(v, nrow = 2, ncol = 2) 
```
Notice the matrix is filled by col by default. We can specify `byrow = TRUE` if we want matrix to be filled by row 
```{r}
matrix(v, nrow = 2, ncol = 2, byrow = T) 
```
We can combine multiple vectors to create matrix either by column or row. Lets define some vectors to see these in action.
```{r}
apple <- c(109.49, 109.90, 109.11, 109.95, 111.03)
ibm <- c(159.82, 160.02, 159.84, 160.35, 164.79)
microsoft  <-  c(59.20, 59.25, 60.22, 59.95, 61.37)
```
We can column bind these vectors together
```{r}
cbind(apple, ibm) 
```
or row bind
```{r}
rbind(apple, ibm) 
```


Similar to vectors, we can also subset Matrix. We can select and/or subset vectors or matrices using `[]` where the first argument is always row and second is always column i.e [row, column].  
```{r}
# column bind all 3 vectors 
stocks <- cbind(apple, ibm, microsoft)
stocks
stocks[2, ] # select 2nd row and all columns
stocks[, 2] # select all rows and 2nd column
stocks[c(1,2), ] # select 1st, 2nd row and all columns
stocks[, c(1,2)] # select all rows and 1st, 2nd column
stocks[c(1,2), c(1,2)] # select 1st, 2nd rows and columns
```

That's all for this blog. If you liked this post or have any suggestions, please let me know. 